---
title: "Formatting and Flagging Example - Discrete Data"
author: "Esther Kennedy"
date: "2023-04-26"
output: pdf_document
---


```{r setup, include=FALSE, results = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Coastal Ocean Acidification Dataset
This data includes surface discrete samples taken from shore by PI Tessa Hill and others. Bottles were immediately assessed for temperature, salinity, and dissolved oxygen using a handheld YSI probe, then preserved with mercuric chloride. All samples were analyzed for either TA and DIC or TA and pH in a lab using established best practices. Note that this dataset also includes field-collected pH data from the YSI probe, but we excluded glass electrode pH data from this compilation due to concerns about its precision and reliability.

This formatting and flagging code is included here as an example of the data manipulation and quality control we engaged in to create our coastal OA synthesis product. In all cases, we erred on the side of leaving data, rather than excluding data. We encourage incestigators using our synthesis dataset to undertake additional QA/QC as necessary for their projects. 


```{r packages, include = FALSE}
# Packages ----------------------------------------------------------------
#Install necessary packages and/or call them
library("tidyverse")
library("lubridate")
library("gridExtra")
library("ggthemes")
```


### Data Formatting
We engaged in basic data wrangling to convert all incoming data to our standard format. While this looked slightly different for each dataset, the basic techniques were the same. 

```{r get_data, results = FALSE, message = FALSE}

# Get the data, make target dataframe------------------------------------------------------------
coast_OA <- read_csv("C:/Users/esthe/Documents/Data Synthesis Project/27_CAOA/coastal_OA_all.csv")
df <- coast_OA
glimpse(df)
summary(df) 


# Light coloumn renaming
df <- df %>% 
  rename(LAT = `LAT DD`, LONG = `LONG DD`, ALK1 = `Alkalinity 1`, ALK2 = `Alkalinity 2`, ALK3 = `Alkalinity 3`, ALK_STD = `Alk Std Dev`, DIC1_umol_L = `DIC 1 (umol/L)`, DIC2_umol_L = `DIC 2 (umol/L)`, DIC_umolkg = `DIC_(umol/kg)`)

#Create a target dataframe
nms <- c("dataset_id","latitude","longitude","depth_m","time_utc","t_C","t_flag",
         "sal_pss","sal_flag","pH_total","pH_flag", "pH_type", "pCO2_uatm","pCO2_flag",
         "pCO2_type","fCO2_uatm","fCO2_flag","fCO2_type","tCO2_umolkg","tCO2_flag",
         "tCO2_type","ta_umolkg","ta_flag","ta_type","do_umolkg","do_sat","do_flag",
         "do_type","chl_ugL","chl_flag","chl_type","si_umolkg","nh4_umolkg","no3_umolkg",
         "no2_umolkg","po3_umolkg","nutr_flag","nutr_type")

target <- as.data.frame(matrix(nrow = nrow(df), ncol = 38))
names(target) <- nms

```

__Copy Data Over and Convert Units__
Changing information to the correct format. Includes and oxygen unit conversion from ppm to % saturation.

```{r units, results = FALSE}
  
#changing timezone to utc
df <- df %>% 
  mutate(date = make_date(year = year, month = month, day = day)) %>% 
  mutate(time_utc = ymd_hms(paste(date, time_pac, sep = " "))) %>% 
  mutate(time_utc = time_utc + hours(8)) %>% 
  dplyr::select(-date)

summary(df)

# Typos in a few Campus point longitudes! Should be 119.8432, but instead they're 199. Fixing:
df<- df %>% 
  mutate(LONG = ifelse(LONG > 190, 119.8432, LONG))


#Easy copy and paste to the target
# Note that we did not retain pH data collected from glass electrode sensors
target <- target %>% 
  mutate(dataset_id = 27) %>%
  mutate(latitude = df$LAT) %>%
  mutate(longitude = -1*df$LONG) %>%   
  mutate(depth_m = 0) %>%
  mutate(t_C = df$Temp) %>%
  mutate(t_flag = 2) %>% 
  mutate(sal_pss = df$Sal) %>%
  mutate(sal_flag = 2) %>% 
  mutate(tCO2_umolkg = df$DIC_umolkg) %>%
  mutate(tCO2_flag = df$DIC_flag) %>% # Original investigators flagged some unreliable samples.
  mutate(tCO2_type = "discrete") %>%
  mutate(ta_umolkg = df$Avg_Alk_meas) %>%
  mutate(ta_flag = df$ta_meas_flag) %>% # Original investigators flagged some unreliable samples.
  mutate(ta_type = "discrete") %>%
  mutate(do_ppm = df$DO) %>% 
  mutate(do_type = "handheld glass electrode")

summary(target)
#Note from 10/2/2021: pre-flagging, there are 140 missing DICs and one missing TA. Since the pH is not good, there could be a maximum of 300 omega calcs from this dataset.

#### Oxygen unit conversion!###
# Convert DO from mg/L to umol/kg
target <- target %>% 
  mutate(do_umolkg = do_ppm*10^3/31.998/1.025)

# Convert the DO to saturation using constants from Benson and Karuse, 1994
source("conversion_functions.R")


target <- target %>% 
  mutate(do_sat = oxy_saturation(t_C, do_umolkg, sal_pss))



#flag all Columbia River data here before we get rid of the site names, 
# As requested by the original investigators
columbia_coords <- df %>% 
  filter(Site_Name == "Columbia River") %>% 
  distinct(LAT, LONG)

target <- target %>% 
  mutate_at(vars(c(ta_flag, tCO2_flag)), ~ ifelse(latitude %in% columbia_coords$LAT & -1*longitude %in% columbia_coords$LONG, 3, .))
#Now there are 165 missing DICs or ones with a 3 flag.

# Reduce down to ideal columns
target <- target %>% 
  dplyr::select(nms)


# Add sample scheme and habitat columns
target <- target %>% 
  mutate(sample_scheme = "intertidal/subtidal hand collection") %>% 
  mutate(habitat = "oceanic")
```
```{r distance_offshore}
# Calculate and add the distance offshore in km.
library("sf")
library("rgeos")
library("sp")
library("raster")

source("calculate_distance.R")
land <- st_read( "americas" )
land <- st_transform(land, crs=4326)

data_sf <- st_as_sf(target, coords = c(x = "longitude", y = "latitude"), crs = 4326)

data_sf[[ 'distance_offshore' ]] <- calculate_distance( data_sf, land, coords="geometry")

target$distance_offshore <- data_sf$distance_offshore

rm(data_sf)
```

### Flagging
For all data sets, our flagging started with the original investigator's notes, if present. In this case, a few samples were flagged as unreliable by the collectors. We will not second-guess those decisions, but will look at all other data to see if more flags need to be applied.

```{r flagging_temp_sal}

target %>% 
  ggplot()+
  geom_point(aes(x = sal_pss, y = t_C, color = t_flag))+
  theme_bw()

target %>% 
  ggplot()+
  geom_point(aes(x = sal_pss, y = t_C, color = sal_flag))+
  theme_bw()
#Look good!

target %>% 
  ggplot()+
  geom_point(aes(x = longitude, y = latitude, color = t_C))+
  theme_bw()

target %>% 
  ggplot()+
  geom_point(aes(x = longitude, y = latitude, color = sal_pss))+
  theme_bw()

target <- target %>% 
  mutate(t_flag = case_when(t_flag == 3 ~ 3,
                            !is.na(t_C) ~ 1)) %>% 
  mutate(sal_flag = case_when(sal_flag == 4 ~ 3,
                              sal_flag == 3 ~ 3,
                              !is.na(sal_pss) ~ 1))

```

```{r do_flagging}
# Note that this is surf-zone dissolved oxygen, so there's unlikely to be much consistency

target %>% 
  ggplot()+
  geom_point(aes(x = t_C, y = do_umolkg, color = sal_pss))+
  theme_bw()
#Buckshot. Nothing stands out.

target %>% 
  ggplot()+
  geom_point(aes(x = do_umolkg, y = do_sat, color = sal_pss))+
  theme_bw()
#Flag the three columbia samples

target <- target %>% 
  mutate(do_flag = case_when(do_flag == 4 ~ 3,
                             do_flag == 3 ~ 3,
                             do_umolkg > 325 & sal_pss < 10 ~ 3,
                             !is.na(do_umolkg) | !is.na(do_sat) ~ 1))

```
```{r flagging_ta_dic}

target %>% 
  ggplot()+
  geom_point(aes(x = ta_umolkg, y = tCO2_umolkg, color = ta_flag))+
  theme_bw()
#Not nice looking, but just a few points to additionally flag

target %>% 
  ggplot()+
  geom_point(aes(x = ta_umolkg, y = tCO2_umolkg, color = tCO2_flag))+
  theme_bw()
#Not nice looking, but just a few points to match with the TA flags

#versus salinity
target %>% 
  ggplot()+
  geom_point(aes(x = sal_pss, y = ta_umolkg, color = ta_flag))+
  theme_bw()

target %>% 
  ggplot()+
  geom_point(aes(x = sal_pss, y = tCO2_umolkg, color = tCO2_flag))+
  theme_bw()

target <- target %>% 
  mutate(ta_flag = case_when(ta_flag == 4 ~ 3,
                             ta_flag == 3 ~ 3,
                             ta_umolkg < 2000 & tCO2_umolkg < 1500 ~ 3,
                             ta_umolkg > 2350 ~ 3,
                             tCO2_umolkg > 2250 ~ 3,
                             !is.na(ta_umolkg) ~ 1)) %>% 
  mutate(tCO2_flag = case_when(tCO2_flag == 4 ~ 3,
                               tCO2_flag == 3 ~ 3,
                               ta_umolkg < 2100 & ta_flag == 3 ~ 3,
                               ta_umolkg > 2350 ~ 3,
                               !is.na(tCO2_umolkg) ~ 1))
```



### Export the data
```{r export}
glimpse(target)

#Make super sure things are in the correct order for later combination
target <- target %>% 
  dplyr::select(nms, sample_scheme, habitat, distance_offshore)

write_csv(target, "C:/Users/esthe/Documents/Data Synthesis Project/27_CAOA/27_final_coastal_OA.csv")


```
